---
title: "Immigration Maps"
author: "Kevin Dyrland"
output:
  html_notebook:
    code_folding: hide
---

# Maps are hard. Prepare to spend at least a day working on this.

## Some thoughts. 

- Do we need to talk about data manipulation?
- We need to talk about where these maps are being used!
- We need to talk about how your audience is going to view the Viz
- Do we want to talk about rmarkdown??
- Maps are hard. gg mapping framework is even harder -- And require an api key?
  - We'll skip `ggmaps()`
  - there are many map frameworks to talk about, we'll do `maps()` and `leaflet()`
- Google early, google often!
- I'm not great at universal design -- many things are implicit in this document that should be explict. ASK ME QUESTIONS and we can update the document together.

```{r packages, message=FALSE}
library(maps)
library(tidyverse)
library(conflicted)
library(geosphere)
library(scales)
library(viridis)
library(leaflet)
library(rgdal)

conflict_prefer("map", "maps")
conflict_prefer("filter", "dplyr")
```

## Connection Map

Let's walk through how to connect points on a world map.

Source: https://www.r-graph-gallery.com/how-to-draw-connecting-routes-on-map-with-r-and-great-circles.html

First we can make some cities:

```{r map_and_cities}
#make some cities (rough long/lat)
Buenos_aires <- c(-58,-34)
Paris <- c(2,49)
Melbourne <- c(145,-38)

# Data frame
data <- rbind(Buenos_aires, Paris, Melbourne) %>% 
  as.data.frame()
colnames(data) <- c("long","lat")

# Show the cities on the map
map('world',
    col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05,
    mar=rep(0,4),border=0, ylim=c(-80,80) 
)
points(x=data$long, y=data$lat, col="slateblue", cex=3, pch=20)
# what happened here?
# we made a dataframe of three cities
# then we used the map() function from the maps package to plot the world
# finally, we added points to the map based on our dataframe
# aside - normally, you can run the plot() function and then the point function separately,
# but for this package they have to be run at the same time. Strange and annoying.
```
Then we'll practice drawing a connecting line...
```{r connection-bad}

# Compute the connection between Melbourne and Buenos Aires 
inter <- gcIntermediate(Buenos_aires, Melbourne, n=50, addStartEnd=TRUE, breakAtDateLine=FALSE)
#what happens if we don't break at the dateline?
# Show this connection
map('world',
    col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05,
    mar=rep(0,4),border=0, ylim=c(-80,80) 
)
points(x=data$long, y=data$lat, col="slateblue", cex=3, pch=20)
lines(inter, col="slateblue", lwd=2)
# what went wrong here?
```
That looks bad! Let's do it a bit better.

```{r connection-better}
plot_my_connection=function( dep_lon, dep_lat, arr_lon, arr_lat, ...){
    #good code has a description right here of a function!
    #basically, we're breaking the matrix in half at the dateline
    #and skipping the pair of points that cross  
    #dep means departure, arr means arrival
    inter <- gcIntermediate(c(dep_lon, dep_lat), c(arr_lon, arr_lat), n=50,
                            addStartEnd=TRUE, breakAtDateLine=F)   
    inter=data.frame(inter)
    diff_of_lon=abs(dep_lon) + abs(arr_lon)
    if(diff_of_lon > 180){
        lines(subset(inter, lon>=0), ...)
        lines(subset(inter, lon<0), ...)
    }else{
        lines(inter, ...)
    }
  
}

map('world',
    col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05,
    mar=rep(0,4),border=0, ylim=c(-80,80)
)
plot_my_connection(Paris[1], Paris[2], Melbourne[1], Melbourne[2], col="slateblue", lwd=2)
plot_my_connection(Buenos_aires[1], Buenos_aires[2], Melbourne[1], Melbourne[2], col="slateblue", lwd=2)
plot_my_connection(Buenos_aires[1], Buenos_aires[2], Paris[1], Paris[2], col="slateblue", lwd=2)
```
Ok, but now let's do some good coding practice. What was annoying about `plot_my_connection()`? How can we improve it? And let's add some color to show the strength of the connection (e.g. the number of immigrants).
```{r connection-best}


#make some data
#head(world.cities)
#get 8 cities
dat.city <- world.cities %>%
  filter(name %in% c("Montreal", "Saint Petersburg", "Paris", "Abidjan", "Salvador",
                     "Buenos Aires", "Nairobi", "Melbourne")
  ) %>%
  filter(country.etc %in% c("Canada", "Russia", "France", "Ivory Coast", "Brazil", 
                            "Argentina", "Kenya", "Australia")
  ) %>%
  filter(!(name == "Paris" & country.etc == "Canada")) #if we don't do this...we get Paris Canada! etc.
#nrow(dat.city)

#I didn't include any in America, China, India. You should pick your own cities!

#make the volume matrix
set.seed(53511)
dat.vol <- dat.city %>%
  expand(sending = name, receiving = name) %>%
  filter(sending != receiving) %>%
  left_join(dat.city, by = c("sending" = "name")) %>%
  left_join(dat.city, by = c("receiving" = "name"), suffix = c(".sending", ".receiving")
  ) %>%
  mutate(volume = sample(1:1000000, nrow(.)))
#expand to get all combinations, drop redundant, then bring lat/long in twice
#(once for sending, once for receiving), then make up volume


vol_lines_thick <- function(data, ...) {
  #condensing plot_my_connection
  #changing line thickness based on relative volume
  city1 <- data %>% select(long.sending,   lat.sending)
  city2 <- data %>% select(long.receiving, lat.receiving)
  inter <- gcIntermediate(city1, city2, n = 50,
                          breakAtDateLine = FALSE, addStartEnd = FALSE) %>%
           as.data.frame()
  diff_of_lon = abs(city1[1]) + abs(city2[1])
  if(diff_of_lon > 180){
        lines(subset(inter, lon>=0), ...)
        lines(subset(inter, lon<0), ...)
    }else{
        lines(inter, ...)
    }
  #holder thought to adjust line thickness
  
  
}
vol_lines_color <- "holder to make function later"

map('world',
    col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05,
    mar=rep(0,4),border=0, ylim=c(-80,80)
)
for(r in 1:nrow(dat.vol)) {
  vol_lines_thick(dat.vol[r,])
}
points(x = dat.city$long, y = dat.city$lat, col="slateblue", cex=2, pch=20)
text(dat.city$name, x=dat.city$long, y=dat.city$lat,  col="slateblue", cex=1, pos=4)

```
Change of plans! Why should I make two functions, one that controls thickness and one that controls color, when I should be able to do it in one? The more generic the function, the better off you are, generally.

Aside - in `ggplot()` you can have colors scale based on your data, auto-magically. Here, we need to make our own colors and put them into the dataset. Source for how to do this: https://stackoverflow.com/questions/45663162/map-values-to-viridis-colours-in-r
```{r connections}
#why am I making two functions?
#just manipulate the data to have the appropiate thickness/color

#color would be simple in ggplot
map_viridis <- function(vec, num) {

  vector_expanded <-round(vec, 1) * 10 # expand to allow for decimal precision
  vector_exp_range <- max(vector_expanded) - min(vector_expanded)

  colour_vector <- viridis(vector_exp_range + 1) # get vector of colour values for all possible decimals between min and max value
  value_to_colour <- colour_vector[num * 10 - min(vector_expanded) + 1] # retrieve colour value for number

  return(value_to_colour)
#https://stackoverflow.com/questions/45663162/map-values-to-viridis-colours-in-r
}

#note - found this while working on the next section - much easier
mypalette2 <- colorNumeric( palette="viridis", domain=dat.vol$volume, na.color="transparent")


dat.vol <- dat.vol %>%
  mutate(thickness = rescale(volume, to = c(1, 5)),
         v.color = map_viridis(dat.vol$volume, volume),
         test.color = mypalette2(volume)
  )

map('world',
    col="#f2f2f2", fill=TRUE, bg="white", lwd=0.05,
    mar=rep(0,4),border=0, ylim=c(-80,80)
)

for(r in 1:nrow(dat.vol)) {
  vol_lines_thick(dat.vol[r,], lwd = dat.vol$thickness[r], col = dat.vol$v.color[r])
}
#do we need to talk about loops?
points(x = dat.city$long, y = dat.city$lat, col="slateblue", cex=2, pch=20)
text(dat.city$name, x=dat.city$long, y=dat.city$lat,  col="slateblue", cex=1, pos=4)

#couple of notes: we need to add a legend! Left to the reader
#because we used fake data, we have St. Pete to Salvador & 
#Salvador to St. Pete - the thicker line dominates. 
# We could probably look into arrows if we wanted...arrows()?
#if time, we should think about how to add arrows showing flow direction
```

## Choropleth 

Use this type when you have a geographic area that you want to color in relation to an ordinal variable.

Source: https://www.r-graph-gallery.com/183-choropleth-map-with-leaflet.html

```{r choro-data}
#commenting out this entire section so it doesn't run again...

# Download the shapefile. (note that I store it in a folder called DATA. You have to change that if needed.)
# download.file("http://thematicmapping.org/downloads/TM_WORLD_BORDERS_SIMPL-0.3.zip", destfile="C:\\Users\\kevin\\Downloads\\DATAworld_shape_file.zip")
# You now have it in your current working directory, have a look!

# Unzip this file. You can do it with R (as below), or clicking on the object you downloaded.
#system("unzip C:\\Users\\kevin\\Downloads\\DATAworld_shape_file.zip")
#  -- > You now have 4 files. One of these files is a .shp file! (TM_WORLD_BORDERS_SIMPL-0.3.shp)
# I was not able to get this to work - I did it manually
```

Simple population Choropleth - not great!
```{r choro-start}
world_spdf <- readOGR( 
  dsn= paste0("C:\\Users\\kevin\\Downloads\\DATAworld_shape_file") , 
  layer="TM_WORLD_BORDERS_SIMPL-0.3",
  verbose=FALSE
)
#this was also a pain

# Clean the data object
world_spdf@data <- world_spdf@data %>%
  mutate(POP2005 = (as.numeric(POP2005) / 1000000) %>% round (2),
         POP2005 = ifelse(POP2005 == 0, NA, POP2005)
  ) 
  
         
#world_spdf@data$POP2005[ which(world_spdf@data$POP2005 == 0)] = NA
#world_spdf@data$POP2005 <- as.numeric(as.character(world_spdf@data$POP2005)) / 1000000 %>% round(2)
#summary(world_spdf@data$POP2005)

#clearly what they did failed somewhere...I think it was the which() command


# Create a color palette for the map:
mypalette <- colorNumeric( palette="viridis", domain=world_spdf@data$POP2005, na.color="transparent")
# mypalette(c(45,43))
# 
# Basic choropleth with leaflet?
leaflet(world_spdf) %>% 
   addTiles()  %>% 
   setView( lat=10, lng=0 , zoom=2) %>%
   addPolygons( fillColor = ~mypalette(POP2005), stroke=FALSE )
 

```

The source says to make a histogram. Always a good idea!

```{r histogram}
world_spdf@data %>%
  ggplot(aes(x = POP2005)) +
  geom_histogram(bins = 20, color = "white")

```
Clearly, a continous distribution is going to be inadequate. Discrete (bins) to the rescue!

```{r final-choropleth}
mybins <- c(0, 10, 20, 50, 100, 500, Inf)
mypalette3 <- colorBin(palette = "viridis", domain = world_spdf@data$POP2005,
                       na.color = "transparent", bins = mybins)
#making a better color function

mytext <- paste0("County: ", world_spdf@data$NAME, "<br/>",
                "Area: ", world_spdf@data$AREA, "<br/>",
                "Population: ", world_spdf$POP2005, " Mil"
          ) %>% #some basic html
          lapply(htmltools::HTML) 
        #allows us to use html...I think we could do this in base r but \n doesn't work

leaflet(world_spdf) %>% 
  addTiles()  %>% 
  setView(lat=10, lng=0 , zoom=2) %>%
  addPolygons( 
    fillColor = ~mypalette3(POP2005), 
    stroke=TRUE, 
    fillOpacity = 0.9, 
    color="white", 
    weight=0.3,
    label = mytext,
    labelOptions = labelOptions( 
      style = list("font-weight" = "normal", padding = "3px 8px"), 
      textsize = "13px", 
      direction = "auto"
    )
  ) %>%
  addLegend( pal=mypalette3, values=~POP2005, opacity=0.9, title = "Population (Mil)", position = "bottomleft" )
```

## What have we learned?

- Shapefiles are hard
- Different maps play with different files better/worse
- Prepare to sink some time into this - this whole presentation take me roughly 4 hours to make and proof and research
- Hopefully, we ran into something unexpected and had to google it

Thanks for your time!
